---
title: "Converting Elan files to all other files"
output: html_notebook
---

# Introduction

This tutorial introduces the ways of converting Elan files to all other files, including Excel (csv and xlsx), and Text.

## Structure of the folders

You should set up three folders in your local computer, including 'InputFiles', 'ProcessingFiles', and 'OutputFiles'. In the folder titled 'OutputFiles', you should have another four folders within it, including 'CSV', 'Excel', 'Text_ALL', and 'Text_PNT'. Please use the exact names for your folders.

## Set your working directory

Before running this script, please set your working directory, which includes all the folders mentioned above.

## Run the script

### Install necessary packages if you don't have them

```{r}
#install these packages if you don't have them
install.packages("writexl")
install.packages("tuneR")
install.packages("parallel")
install.packages("audio")
install.packages("xml2")
```

### Install the package 'elan' as it's a tricky one

```{r}
#run the following lines to get the 'elan' package
library(devtools)
install_github("dalejbarr/elan")
```

### Load all your packages

```{r}
#load all the necesary packages
library(writexl)
library(tuneR)
library(parallel)
library(audio)
library(elan)
library(xml2)
```

### Run the script

```{r}
options(scipen=999)

convert_duration = function(duration){
  duration = as.numeric(duration)
  duration_str = as.character(duration)
  #------------------------------------
  
  #mills = as.numeric(gsub(".*\\.","",format(round(duration, 3), nsmall = 3)))
  mills = gsub(".*\\.","",format(round(duration, 3), nsmall = 3))
  
  #------------------------------------
  secs = floor(duration%%60)
  if(secs < 10){
    secs = paste0('0', floor(secs))
  }else{
    secs = floor(secs)
  }
  #------------------------------------
  mins = floor((duration/60)%%60)
  if(mins < 10){
    mins = paste0('0', floor(mins))
  }else{
    mins = floor(mins)
  }
  
  #------------------------------------
  hrs = floor((duration/(60*60))%%24)
  if(hrs < 10){
    hrs = paste0('0', floor(hrs))
  }else{
    hrs = floor(hrs)
  }
  tmp_time = paste0(hrs, ':', mins, ':', secs, '.', mills)
  
  return(tmp_time)
}

convert_to_msecs = function(mills){
  msecs_digits = nchar(mills) - 3
  msecs_pattern = paste0('^([0-9]{', msecs_digits, '})([0-9]+)$')
  split_msecs = gsub(msecs_pattern, '\\1.\\2', mills)
  return(split_msecs)
}

specify_decimal <- function(x) format(round(x, 3), nsmall=3)

#======================================================================
#======================================================================

individual_columns = unlist(strsplit('PNT INT SP ENV', ' '))
contiguous_columns = unlist(strsplit("Quotatives Me_possessive There's Social_info", ' '))
keep_col_names = unlist(strsplit("PNT INT SP ENV Anonymity Quotatives Me_possessive There's Social_info", ' '))

SS_eaf2df <- function(eaf_file) {
  
  final_df_col_names = c('Transcript', 'IU_number', 'start_time', 'end_time', 'speaker', 'IU', 'start_time_mills', 'end_time_mills')
  final_df = data.frame(matrix(nrow = 0, ncol = length(final_df_col_names)))
  names(final_df) = final_df_col_names
  
  #eaf_dat <- read_eaf(eaf_file)
  eaf_dat <- efileAnnotations(eaf_file)
  #eaf_dat$EXT_REF <- NULL
  
  #eaf_dat <- elanTree(eaf_file)
  
  #get names of tiers
  eaf_tiers_all <- unique(eaf_dat$TIER_ID)
  
  eaf_tiers_all = eaf_tiers_all[!grepl(paste(c("Numbers", "Word_list", "LaBB", "Extra transcription section", "Powerhouse", "traini", "transcription"), collapse = "|"), eaf_tiers_all)]
  
  except_tiers = '!Numbers|!Word_list|!LaBB|!Extra transcription section|!Powerhouse|!traini|!transcription'
  
  eaf_dat$t0m <- as.numeric(unlist(lapply(eaf_dat$t0, convert_to_msecs)))
  eaf_dat$t1m <- as.numeric(unlist(lapply(eaf_dat$t1, convert_to_msecs)))
  eaf_dat$t0tm <- unlist(lapply(eaf_dat$t0m, convert_duration))
  eaf_dat$t1tm <- unlist(lapply(eaf_dat$t1m, convert_duration))
  
  for(tier_i in keep_col_names){
    
    #extract the tier
    eaf_tiers <- eaf_dat[grepl(paste0('^', tier_i), eaf_dat$TIER_ID),]
    all_speakers <- unique(eaf_tiers$TIER_ID)
    tmp_eaf_tiers_length = length(all_speakers)
    
    if(tmp_eaf_tiers_length > 0){
      for(eaf_tiers_i in 1:tmp_eaf_tiers_length){
        
        #full tier name
        speaker_tier <- all_speakers[eaf_tiers_i]
        
        if(length(speaker_tier) > 0){
          
          loop_final_df <- eaf_tiers[grepl(paste0('^', speaker_tier),  eaf_tiers$TIER_ID),]
          names(loop_final_df) <- unlist(strsplit('speaker ANNOTATION_ID IU t0.id t0 t1.id t1 Transcript atype start_time_mills end_time_mills start_time end_time', ' '))
          loop_final_df$IU_number = 1:nrow(loop_final_df)
          loop_final_df = loop_final_df[,c('Transcript', 'IU_number', 'start_time', 'end_time', 'speaker', 'IU', 'start_time_mills', 'end_time_mills')]
          loop_final_df$Transcript = gsub("./InputFiles/|.eaf", "", eaf_file)
          
          final_df = rbind(final_df, loop_final_df)
        }
        
      }
    }
  }
  
  #fix quotatives
  tmpq <- final_df[final_df$speaker == 'Quotatives',]
  tmppnt <- final_df[grep('PNT', final_df$speaker),]
  
  if(nrow(tmpq) > 0){
    
    #start time
    for(qi in 1:nrow(tmpq)){
      
      tmpl <- tmpq[qi,]
      
      # View(tmpl)
      # View(tmppnt)
      closestValueIndex <- which(abs(tmppnt$start_time_mills-tmpl$start_time_mills)==
                                   min(abs(tmppnt$start_time_mills-tmpl$start_time_mills)))
      closestValue <- tmppnt$start_time_mills[closestValueIndex]
      
      tmpdiff <- abs(closestValue - tmpl$start_time_mills)
      
      if(tmpdiff < 0.010){
        final_df[final_df$IU_number == tmpq$IU_number[qi] &
                   final_df$speaker == 'Quotatives', 'start_time_mills'] <- closestValue
      }
      
    }
    
    #end time
    for(qi in 1:nrow(tmpq)){
      
      tmpl <- tmpq[qi,]
      
      closestValueIndex <- which(abs(tmppnt$end_time_mills-tmpl$end_time_mills)==
                                   min(abs(tmppnt$end_time_mills-tmpl$end_time_mills)))
      closestValue <- tmppnt$end_time_mills[closestValueIndex]
      
      tmpdiff <- abs(closestValue - tmpl$end_time_mills)
      
      if(tmpdiff < 0.010){
        final_df[final_df$IU_number == tmpq$IU_number[qi] &
                   final_df$speaker == 'Quotatives', 'end_time_mills'] <- closestValue
      }
      
    }
  }
  
  return(final_df)
}   


in_files = dir("./InputFiles", pattern = ".eaf", full.names = TRUE)
file_elan_names = c('Transcript', 'IU_number', 'start_time', 'end_time', 'speaker', 'IU', 'start_time_mills', 'end_time_mills',
                    "Anonymity","Social_info","Linguistic_info","Quotatives","Me_possessive","There's","Social_info")
final_df_elan = data.frame(matrix(nrow = 0, ncol = length(file_elan_names)))
names(final_df_elan) = file_elan_names

for(i in in_files){
  print(i)
  
  #imports the Elan file and converts it to a data.frame
  tmp_df = SS_eaf2df(i)
  
  #organises the IUs based on initial time
  tmp_df = tmp_df[with(tmp_df, order(start_time)), ]
  
  #creates a column with the number ofm IUs
  tmp_df$IU_number = 1:nrow(tmp_df)
  
  #tmp_df$sum <- tmp_df$end_time_mills - tmp_df$start_time_mills
  
  #extra tiers whose information is extracted and attached to the corresponding speaker
  extra_tiers = c("Anonymity","Linguistic_info","Quotatives","Me_possessive","There's","Social_info")
  #extra_tiers = c("Anonymity","Social_info","Linguistic_info","Quotative","Me_possessive","There's")
  
  #creates respective columns of extra tiers in the data.frame
  for(extra_i in extra_tiers){
    tmp_df[[extra_i]] = ""
  }
  
  #creates atemporal data.frame with ONLY the extra tiers specified for extracting values
  df_noExtra = tmp_df[!tmp_df$speaker %in% extra_tiers,]
  df_noExtra$ordered = 1:nrow(df_noExtra)
  
  for(speaker_i in individual_columns){
    #gets the names of the specified speaker
    tmp_speaker_tier = unique(df_noExtra$speaker)[grepl(speaker_i, unique(df_noExtra$speaker))]
    
    #iterates for each extra tier
    for(j in extra_tiers){
      
      #if the extra tier exists in the data frame
      if(length(which(unique(tmp_df$speaker) == j)) != 0)
      {
        # if(j == "Anonymity" & speaker_i == 'INT'){
        #   sfsgsg
        # }
        #stores the iterated extra tier
        tmp_extra_tier = j
        #the speaker tier PLUS the extra tier
        tmp_tiers = c(tmp_speaker_tier, tmp_extra_tier)
        #data frame with the speaker tier PLUS the extra tier
        main_df = tmp_df[tmp_df$speaker %in% tmp_tiers,]
        
        #delete all empty IUs
        main_df = main_df[main_df$IU != '',]
        
        #initital time of the data frame with speaker
        init_main = main_df[main_df$speaker == tmp_speaker_tier, 'start_time_mills']
        #final time of the data frame with speaker
        end_main = main_df[main_df$speaker== tmp_speaker_tier, 'end_time_mills']
        
        #initital time of the data frame with extra tier
        init_extra = main_df[main_df$speaker == tmp_extra_tier, 'start_time_mills']
        #final time of the data frame with extra tier
        end_extra = main_df[main_df$speaker == tmp_extra_tier, 'end_time_mills']
        
        tmp_main_df = main_df[main_df$speaker == tmp_speaker_tier,]
        
        #iterates for EACH extra tier interval
        for(k in 1:length(init_extra)){
          
          #locates the start of all the intervals lower than the initial to find
          tmp_lower_index_all = which(tmp_main_df[,'start_time_mills'] <= init_extra[k])
          #tmp_lower_index = tmp_lower_index_all[length(tmp_lower_index_all)]
          
          if(length(tmp_lower_index_all) != 0){
            #locates the row index of the first IU lower than the extra tier
            tmp_lower_index = tmp_lower_index_all[tmp_lower_index_all[length(tmp_lower_index_all)]]
            
            #locates the row index of all the IUs higher than the extra tier
            tmp_higher_index_all = which(tmp_main_df[,'end_time_mills'] >= end_extra[k])
            
            #if the first higer index is not equal to zero and the length of the lower index is not equal to zero
            if(length(tmp_higher_index_all[1]) != 0 & length(tmp_lower_index) != 0){
              
              #if any of this is not NA
              if(!is.na(tmp_higher_index_all[1]) & !is.na(tmp_lower_index)){
                #sets the higer index
                
                #if the first index is larger than the lower index
                if(tmp_higher_index_all[1] > tmp_lower_index){
                  
                  #check whether the upper boundary is inside an existing IU in the same speaker
                  if(tmp_main_df[tmp_higher_index_all[1],'start_time_mills'] <= end_extra[k] &&
                     tmp_main_df[tmp_higher_index_all[1],'start_time_mills'] <= init_extra[k]){
                    if(abs(end_extra[k] - tmp_main_df[tmp_higher_index_all[1],'end_time_mills']) == 0.001 ){
                      tmp_higher_index = tmp_higher_index_all[1] - 1
                    }else{
                      tmp_higher_index = tmp_higher_index_all[1]
                    }
                  }
                }else{
                  tmp_higher_index = tmp_higher_index_all[1]
                }
                
                #check whether the upper boundary is inside an existing IU in the same speaker
                if(tmp_main_df[tmp_higher_index_all[1],'start_time_mills'] <= end_extra[k] &&
                   tmp_main_df[tmp_higher_index_all[1],'start_time_mills'] <= init_extra[k]){
                  
                  #the initial time in the speaker tier - data frame with ONLY speaker and extra tier
                  initial_time = tmp_main_df[tmp_lower_index, 'start_time_mills']
                  #the final time the speaker tier - data frame with ONLY speaker and extra tier
                  final_time = tmp_main_df[tmp_higher_index, 'end_time_mills']
                  
                  #locates the IU interval in the main data frame - initial time
                  one_init_lower = main_df[main_df$speaker == tmp_speaker_tier & main_df$start_time_mills == initial_time, 'IU_number']
                  #locates the IU interval in the main data frame - final time
                  one_init_upper = main_df[main_df$speaker == tmp_speaker_tier & main_df$end_time_mills == final_time, 'IU_number']
                  
                  #gets the row number of the speaker in which the extra tier interval should be attached - lower value
                  lower_index = df_noExtra[df_noExtra$IU_number == one_init_lower, 'ordered']
                  #gets the row number of the speaker in which the extra tier interval should be attached - upper value
                  upper_index = df_noExtra[df_noExtra$IU_number == one_init_upper, 'ordered']
                  
                  #gets the label of the extra tier to be attached to the speaker IU
                  tmp_extra_label = main_df[main_df$speaker == tmp_extra_tier & main_df$start_time_mills == init_extra[k], 'IU']
                  
                  #gets the text of the field
                  tmp_existing_label = df_noExtra[lower_index,j]
                  
                  # if(j == "Quotatives" & k  == 5){
                  #   asdfrf
                  # }
                  
                  #if there is no previous label, the new label is added
                  #if there is a label already, the new label is attached
                  if(tmp_existing_label == ""){
                    df_noExtra[lower_index:upper_index,j] = tmp_extra_label
                  }else{
                    df_noExtra[lower_index:upper_index,j] = paste(tmp_existing_label, tmp_extra_label, sep=',')
                  }
                  
                }
                
              }
            }
          }
          
        }
        
      }
    }
    
  }
  
  df_noExtra$IU_number = 1:nrow(df_noExtra)
  
  tmp_df_save = df_noExtra
  tmp_df_save$ordered = NULL
  tmp_df_save$start_time_mills = NULL
  tmp_df_save$end_time_mills = NULL
  rownames(tmp_df_save) <- NULL
  
  #ms saving
  tmp_df_save_ms = tmp_df_save
  tmp_name_ms = paste0("./OutputFiles/CSV/", gsub("./InputFiles/|.eaf", "", i), '.csv')
  tmp_name_xlsx_ms = paste0("./OutputFiles/Excel/", gsub("./InputFiles/|.eaf", "", i), '.xlsx')
  
  # write.xlsx(tmp_df_save_ms, tmp_name_xlsx_ms, sheetName="Sheet1", col.names=TRUE, row.names=FALSE, append=FALSE)
  write_xlsx(tmp_df_save_ms, tmp_name_xlsx_ms, col_names=TRUE)
  write.csv(tmp_df_save_ms, tmp_name_ms)
  
  tmp_name_tab = paste0("./OutputFiles/TEXT_All/", gsub("./InputFiles/|.eaf", "", i), '.txt')
  tmp_name_tab_PNT = paste0("./OutputFiles/Text_PNT/", gsub("./InputFiles/|.eaf", "", i), '.txt')
  
  #tab file
  tdtf = tmp_df_save_ms[,c('start_time', 'speaker', 'IU')]
  tdtf = tdtf[grepl(paste(individual_columns, collapse="|"), tdtf$speaker),]
  
  write.table(tdtf, tmp_name_tab, sep="\t", quote = F, row.names = F, col.names = F)
  
  #only PNT
  txtin = tdtf[grep('^PNT', tdtf$speaker),]
  txtout = txtin[,c('start_time', 'speaker', 'IU')]
  write.table(txtout, tmp_name_tab_PNT, sep="\t", row.names = F, col.names = F, quote = F)
  
  final_df_elan = rbind(final_df_elan, df_noExtra)
}

final_df_elan$start_time_mills = NULL
final_df_elan$end_time_mills = NULL
write.csv(final_df_elan, './ProcessingFiles/all_files.csv')

warnings()
```
